OA Share

### 2021-01-27 analysis/manuscript.Rmd
tmp save

@@

During the same period, Elsevier published `r all_oa` OA articles in total (i.e., in hybrid and full OA journals), representing `r all_oa / all_els * 100`% of the total article volume (n=`r all_els`). Looking at Figure \ref{fig:oa_volume_fig}, it is apparent that the OA article volume of hybrid journals (n=`r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(n)`, `r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(prop) * 100`%) lagged behind that of Elsevier's Open Archive program (n=`r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(n)`, `r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(prop) * 100`%) and full OA journals (n=`r filter(summary_oa_type, type == "full_oa") %>% pull(n)`, `r filter(summary_oa_type, type == "full_oa") %>% pull(prop) * 100`%), which included `r sum(mirror_jn$all_articles)` articles published in `r length(unique(mirror_jn$container_title))` mirror journals.

@@

As far as we observed, OA articles in Elsevier hybrid journals were published under two possible CC licenses--CC BY or CC BY-NC-ND. CC BY allows others to distribute, remix, adapt, and build upon the licensed work, including for commercial purposes, as long as the original author is credited. In contrast, CC BY-NC-ND is less permissive as it prohibits commercial reuse and derivatives. In the five-year period from 2015-2019, the proportion of hybrid OA articles published under a CC BY-NC-ND license marginally increased and maintained a greater share than that of CC BY (see Figure \ref{fig:oa_volume_fig}-B). Interestingly, Figure \ref{fig:oa_volume_fig}-B also reveals that hybrid journals had the highest number and proportion of openly available articles licensed under CC BY (n=`r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(n)`, `r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(prop) *100`%) compared to full OA journals  (n=`r filter(cc_by_by_type, type == "full_oa") %>% pull(n)`, `r filter(cc_by_by_type, type == "full_oa") %>% pull(prop) *100`%) and Elsevier's delayed OA program (n=`r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(n)`, `r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(prop) *100`%) . Most  delayed OA articles were provided under an Elsevier user license (Els-User), which prohibits reuse for commercial purposes.

@@

Figure \ref{fig:sponsor_sector} shows the yearly distribution of central invoiced articles by year and institutional sector. While UK-based open access invoices were mainly addressed to discipline-specific governmental and non-profit research funders,  the situation is different in the Netherlands and Norway where national consortia of institutions belonging to the sector of higher education (e.g. universities) had transformative agreements with Elsevier. In 2019, Elsevier also launched similar agreements in other countries with lower publication output including Hungary and Poland. Besides, we found that invoiced funding bodies from the UK and the US mainly represent discipline-specific funders, while the remaining funding bodies from other countries are focused on a broad variety of disciplines (see Table \ref{tab:subject_profile}).

@@

Table \ref{tab:sponsor_top_10} presents the top ten of `r hybrid_invoice %>% filter(!is.na(oa_sponsor_cleaned)) %>% distinct(oa_sponsor_cleaned) %>% nrow()` invoiced institutions and organizations in our sample. Together, they accounted for around 80% of centrally invoiced APCs. The table also highlights the number of distinct journals and the share of CC BY licensed articles. Notably, UK-based research funders and the US-based Bill and Melinda Gates Foundation mainly sponsored articles published under a CC BY license. In contrast, articles funded by VSNU, the European Research Council, and Norway Institutes had a much lower proportion of CC BY.

@@

ind_sponsor_df %>%
  bind_rows(overall_ind_sponsor_df) %>%
  mutate(oa_sponsor = gsub("&", "and", oa_sponsor)) %>%
  mutate_at(4:6, function(x){
    round(x*100,0)
  }) %>%
  mutate_at(5:6, function(x){
    cell_spec(x, background = make_pal(x))}) %>%
  kbl("latex", booktabs = TRUE, escape = FALSE,
      col.names = c("Hybrid OA sponsors", "Journals", "Articles", "\\%", "CC BY", "OAPC"),
      align = c("lrrrrr"),
      linesep = "",
      caption = "Top 10 Elsevier hybrid OA sponsors by centrally invoiced articles 2015-2019.  The table furthermore presents the number of Elsevier hybrid journals with at least one invoiced article and two indicators of compliance: proportion of articles provided under a CC BY license and where spending was disclosed through the Open APC Initiative (OAPC). VSNU and Norway Institutes paid for OA through transformative agreements according to the ESAC Transformative Agreement Registry.") %>%
  row_spec(11:12, hline_after = TRUE) %>%
  add_header_above(c(" " = 4, "Compliance (in %)" = 2)) %>%
   kable_styling(latex_options = c("scale_down"))
```

The table also highlights the proportion of APC funding publicly disclosed through the Open APC initiative, suggesting that funders with a large CC BY share were also more willing to share their spending publicly. On the other hand, APC funding activities through transformative agreements were not publicly shared.

Figure \ref{fig:field_funder_matrix} shows the subject profile of OA funders relative to the journals’ subject fields. As expected, the figure shows that subject-specific funders tended to focus on a limited set of subject fields. In contrast, national consortia, such as VSNU and Norwegian Institutes (NO), financed OA articles  for a broader range of fields. However, within these consortia, the subject distribution also varied. VSNU OA spending, for example, had a strong focus on articles relating to Medicine, Environmental Science, and Energy, presumably reflecting disciplines with generally strong publication output at Dutch universities. Norwegian universities mostly published hybrid OA in Engineering, Environmental Science and Agricultural and Biological Sciences journals.

>>>>

# Background

The ability to identify and discover hybrid OA articles plays a fundamental role in monitoring hybrid OA, yet time and again, previous studies have underscored the difficulties involved in this (cf. [@Akbaritabar_2019; @van_Leeuwen_2018; @Laakso_2016; @Emery_2013; @Bj_rk_2012]. Insights into the volume of hybrid OA and the various levels and sources of APC funding could create a better understanding of the costs involved in scholarly publishing and, thus, strengthen institutions' bargaining power in negotiations with publishers, and promote greater equity in scholarly publishing. In the following section, we examine how previous research has tackled the challenge of identifying hybrid OA content and what these studies taught us about the uptake and funding sources of hybrid OA.

## Prevalence / uptake studies

Since there is no standardized way to measure the uptake of hybrid OA, estimates can vary greatly depending on how uptake is operationalized-—for example, through the number of hybrid journals with at least one OA article, the average number of OA articles per hybrid journal, or the share of OA articles compared to subscription articles.

In the first article-level study on hybrid OA, @Laakso_2016 reported that between 2007 and 2013, the five largest publishers--Elsevier, Sage, Springer, Taylor & Francis, and Wiley--recorded growing numbers of hybrid journals that coincided with a twenty-fold increase in hybrid OA articles (from 666 to 13,994). As a result, hybrid journals with at least one OA article more than doubled in number (1,082 in 2009, 2,714 in 2013) but fell by 31% relative to the number of hybrid journals in total. Although Laakso and Björk's (2016) presented a comprehensive overview tracking the growth of hybrid OA over time and across multiple publishers, their exploratory approach is unsuitable for repeated use as it required a considerable amount of manual data collection and cleaning. Minimizing such manual tasks, @Nelson_2017 narrowed their focus to one publisher--the American Chemical Society (ACS)--and requested a list of all hybrid OA articles published from 2006-2011 directly from the ACS. To assess the hybrid OA uptake, the authors aggregated and compared the number of OA (n=814) and subscription articles (n=27,621), finding that over the five years, 2.9% of ACS articles were published as hybrid OA.

Contrary to the previous two studies, @Kirkman_2018 took a research funder as a starting point, reporting a 26% share of hybrid OA for articles funded by the Australian National Health and Medical Research Council from 2012-2014 (816 of 3,190). Kirkman collected research articles using the Funding Agency and Funding Text search fields on Web of Science, and classified articles as OA when they were freely available on the publisher website, and further distinguished hybrid OA articles based on journal-level information from the publisher website, such as hybrid OA journal and APC lists. Another approach is presented by @P_l_nen_2020, who used current research information system (CRIS) data—institutional publication data—to study the extent of OA among Finnish research publications from 2016-2017 (n=34,507 research articles). As part of the analysis, the authors identified hybrid OA articles through a dedicated OA status metadata field that has been mandatory to report in Finland since 2016 and found a 7% share of hybrid OA. However, the drawbacks of Kirkman's (2018) and Pölönen et al.'s (2020) approaches are that such data sources might not be openly available or, regarding CRIS data, might not be available at all since not all countries maintain comprehensive institutional publication data.

Other studies assessed the share of (hybrid) OA for the entire corpus of scholarly research articles. For instance, using research articles with an assigned DOI and indexed in the Web of Science Core Collection (n=2,610,305) as a benchmark, @Mart_n_Mart_n_2018 assessed how many research articles from 2009 and 2014 are freely available through Google Scholar. After identifying OA articles through licensing information from Crossref, the authors estimated the share of hybrid OA at around 0.5% in 2009 and 1.5% in 2014. In a similar vein, @Piwowar_2018 drew on Crossref to identify content licenses. The authors followed a data-driven approach based on Unpaywall data, which is openly available and contained 66 million Crossref DOIs at the time, to determine how much of the scholarly literature is OA. Based on a random sample of 100,000 Crossref DOIs, the authors found 3.6% of hybrid OA, while pointing towards a growing trend in recent years that recorded a hybrid OA share of 9.4% of all articles published in 2015. While Piwowar et al.'s (2018) and Martín-Martín et al.'s (2018) methods are well-suited for large-scale and repeated analyses of the prevalence of hybrid OA, their main limitation is that they lack information about publishing costs.

Moreover, Unpaywall's hybrid OA detection method has been refined several times since its launch in 2017, reflecting growing albeit incomprehensive license information in Crossref metadata. To overcome these shortcomings, the developers extended previous web-scraping approaches to better distinguish between immediate and delayed OA based on licenses at the date of observation @Piwowar_2019. Relatedly, in October 2020, Elsevier's license for delayed OA content was added to Unpaywall's list of license patterns [@Unpaywall_oa_license]. Finally, at the journal-level, a more fine-grained identification of full OA journals using OA article information was developed to better distinguish them from subscription journals including hybrid journals [@Unpaywall_types]. Because of these ongoing changes due to non-standardized data, uptake studies reporting hybrid OA uptake based on Unpaywall data cannot easily be compared with each other.

## Financial studies

Some countries like Sweden and the United Kingdom (UK) have negotiated national publishing agreements that allow authors to publish hybrid OA at a discounted fee or for free. These agreements can serve as the basis for studies on the uptake of hybrid OA or cost-benefit analyses (cf. @Olsson1271866; @Lawson_2018) but are often limited in scope as they tend to focus on a single country and rely on data that seldom is publicly available. Hence, while being valuable for specific use-cases, such studies can neither provide a comprehensive understanding of how the scholarly publishing system functions beyond national boundaries nor facilitate collective sharing of financial information about hybrid OA.

Pinfield et al.'s (2016) analysis of APC payment records provided by 23 UK higher education institutions revealed a sharp increase in central payments from 2007-2013, which was largely attributed to the introduction of block grants by Research Councils UK (RCUK) and non-compliance sanctions by the Wellcome Trust. Moreover, the study showed that OA fees were paid almost exclusively through RCUK and Wellcome block grants (92%), and only a small number of APCs were paid through internal funding (7%). In contrast, a recent Springer Nature survey found that authors draw on a range of funding sources to cover OA fees, such as dedicated institutional OA funds, block grants, OA agreements, or research grants [@Monaghan_2020]. Most hybrid OA authors were supported through dedicated institutional OA funds (43%, excluding block grants) and OA agreements with Springer Nature (34%). The differences between these two studies may be due to changes in policy and funding arrangements—in particular, the introduction of OA agreements—since @Pinfield_2016, or due to Monaghan et al.'s (2020) more regionally diverse sample, indicating regional differences in APC funding.

Regional and policy differences in APC payments also came to light in a study by @Jahn_2016 that analyzed APC records from 30 German higher education and research institutions crowd-sourced by the Open APC initiative, the Austrian Science Fund (FWF), Jisc, and the Wellcome Trust. In particular, the study revealed large differences in the amount of hybrid OA funded from 2014-2015. Whilst hybrid OA accounted for less than 1% of APCs paid by German institutions (23 of 3,846), the three non-German research funders recorded a hybrid share of75% (11,533 of 15,779). According to @Jahn_2016, this could point towards differences in science policy, such as hybrid OA being supported by the three non-German research funders but not by Germany's largest national funder (DFG). Another possibility is that German hybrid OA fees were paid from budgets not reported to the Open APC initiative, from where the authors acquired the German APC data. Among these unreported funds are research grants and research unit budgets, which author surveys identified  as funding sources for hybrid OA fees [@Monaghan_2020; @Graaf_2017]. As such, Jahn and Tullney's (2016) findings could reflect the complexities and potential limitations of institutional OA spending data that Pinfield et al. (2016) and Monaghan et al. (2020) attributed to incomplete or missing records.

As this brief review has shown, the most comprehensive studies approach hybrid OA funding from an institutional perspective that focuses on research institutions or funders and draws on internal records or crowd-sourced data like from the Open APC initiative. The advantages of Open APC data are that they reflect the actual APCs paid rather than list prices; they are not limited to certain publishers or journals; they are provided in a standardized format regardless of the publisher. On the other hand, the drawbacks include the limited number of participating institutions that are almost exclusively based in Europe and the dependency on participating organizations for data integrity. Moreover, since Open APC data does not include waivers or free APCs, it cannot present a complete picture of all publishing activities and expenses. This, however, would pose less of an issue when analyzing just hybrid OA.

## Data-driven workflows for hybrid OA

In September 2018, coalition S presented a set of principles known as Plan S and announced. that from 2021 onwards, signatories would no longer fund publications in hybrid OA journals. This principle includes so-called "mirror" and sister journals—hybrid journals with fully OA counterparts or split into their fully OA and subscription parts—that have launched since Plan S was announced but makes an exception for hybrid journals with transformative agreements [@Frantsv_g_2019]. While the scale and means of these transformative agreements may differ, many address the lack of transparency surrounding hybrid OA by making publisher-provided metadata a defining element of publishing workflows between institutions and publishers [@Geschuhn_2017; @Marques_2019]

Elsevier, one of the world's largest scholarly publishers, provides article-level metadata to Crossref, including license information, which can be used to identify OA articles, and links to HTML and XML full-text versions for text and data mining purposes. Surprisingly, besides the full-text article, the XML files also contain invoicing information and the article's OA status. Table \ref{tab:els_xml} summarizes the key metadata elements, which comprise the type of entity to which the APC was invoiced (`openaccessSponsorType`) and the name of that entity if available (`openaccessSponsorName`). Moreover, the article-level metadata indicates if OA was provided through Elsevier's Open Archive program (`openArchiveArticle`), which enables free access to participating subscription journals after an embargo period (i.e., delayed OA; @Els_Archive). As we have been unable to identify similar efforts by other large publishers (e.g., Springer Nature), this increased transparency through openly available bibliometric and financial data at the article level seems unique to Elsevier.

\begin{table}[H]
\caption{\label{tab:els_xml}Overview of hybrid open access (OA) uptake across Elsevier hybrid journals with at least one OA article 2015-2019.}
\centering
\begin{tabular}[t]{lp{10cm}}
\toprule
XML Node & Description\\
\midrule
\texttt{openaccessArticle} & Indicates if the article is open
access. \\
\texttt{openaccessSponsorType} & Invoice recipient type.\\
\texttt{openaccessSponsorName} & Entity that received Elsevier's APC
invoice.\\
\texttt{openArchiveArticle} & Indicates if open access was provided through
Elsevier's Open Archive programme.\tabularnewline \\
\bottomrule
\end{tabular}
\end{table}

Table 1: Overview of metadata provided in XML-based Elsevier full-texts representing information about open access status and invoicing. A sample XML is available here: https://api.elsevier.com/content/article/PII:S0169409X18301479?httpAccept=text/xml

In this study, we utilize Elsevier's publicly available metadata to analyze the uptake and invoicing of their hybrid OA publishing option. In particular, we use this novel methodology to answer the following questions:

- What was the uptake of Elsevier's hybrid OA publishing option between 2015 and 2019?

- Through which channels were hybrid APCs invoiced, and who were the recipients?

# Methodology

For this study, we collected data relating to Elsevier's hybrid OA option by drawing on multiple freely available data sources. We identified Elsevier hybrid journals through Elsevier’s APC list and supplemented our sample with Crossref metadata and text-mined invoicing data to investigate the financing of immediate OA articles provided under a Creative Commons license in subscription-based journals. Figure \ref{fig:workflow} visualizes the automated workflow we used to collect data from Elsevier and Crossref.

{\centering \includegraphics[width=0.95\textwidth]{../figure/els_flow.pdf}

\caption{Data Analytics Workflow to obtain article-level OA and invoicing data using freely available data sources provided by Elsevier and Crossref.
}\label{fig:workflow}
\end{figure}

First, we obtained hybrid journals through an Elsevier APC price list from May 2020 provided as captured by Matthias (2020). We only included hybrid journals that were published under the hybrid model throughout the entire period of our analysis. Hence, we excluded journals that flipped from OA to hybrid and from hybrid to OA. This approach allowed us to ensure that the article-level differences we observed were not due changes at the journal-level (i.e., increases or decreases in number). Overall, we identified 1,970 unique hybrid journals.

Next, we used an openly available Crossref database snapshot [@Crossref_2020], which contains all Crossref records registered until March 2020, to calculate the journals' combined article volume for the five-year period from 2015-2019. We only included articles published in regular issues aside from supplements containing conference contributions like meeting abstracts, indicated by non-numeric pagination. We excluded non-scholarly journal content, such as the table of contents and lists of reviewers, following Unpaywall's paratext recognition approach [@Unpaywall_para], which we expanded to include patterns indicating corrections. Furthermore, we categorized the articles by subject according to the All Science Journal Classification code (ASJC) and added citation indicators from the Scopus Source Title list (version June 2020) based on the journal they were published in.

Crossref metadata also contains license information that indicates if and under which terms the publisher has made articles openly available on their website [@Hendricks_2020]. We identified OA articles through Creative Commons (CC) licenses and then downloaded the XML version of all CC-licensed articles published in a hybrid journal. From the XML files, we determined whether or not the articles were immediate OA using the XML node `openArchiveArticle` and measured the uptake of hybrid OA through three bibliometric indicators [@Nelson_2017; @Laakso_2016]-—the number of hybrid journals with at least one OA article, the number of hybrid OA articles, and the number of hybrid OA articles relative to the number of closed-accessed articles in hybrid journals with at least one OA article.

Moreover, we obtained APC invoicing metadata from the XML (see Table 1), which enabled us to identify the invoice channels and recipients of hybrid OA APCs. Based on the `openaccessSponsorType` node in the XML files, we distinguish between four invoice channels, including invoices billed to the corresponding author, as part of publishing agreements with funding bodies (hereinafter "agreements"; cf. @Els_Agreements), fee waivers (e.g., in "cases of genuine need" or due to society or university sponsorships, cf. ref), and other types not specified by Elsevier. Finally, when hybrid APCs were invoiced as part of OA agreements, we identified invoice recipients through the `openaccessSponsorName` node.

We manually classified invoice recipients based on their institutional sectors, countries, and primary research areas. Following the OECD's Frascati Manual [@Frascati, p. 91], we coded for four sectors, business enterprise, government, higher education, and private non-profit. The government sector consists of federal, state, and municipal governments, plus non-profit institutions controlled by government units [@Frascati, p. 100]. The higher education sector comprises institutions that perform research and development and provide formal tertiary education services, such as universities or research institutes [@Frascati, p. 101]. Legal or social entities that produce goods and services without generating profit, such as learned societies or charities, constitute the private non-profit sector [@Frascati, p. 288]. Due to the low article volume, we combined invoice recipients Elsevier listed as "authors" and "third-party sponsor", and those we had coded as business enterprises [@Frascati, p. 97], into "Others".  Moreover, we categorized invoice recipients according to the country or countries representing their scope of funding and based on the following primary research areas health sciences, life sciences, physical sciences and mathematics, social sciences and humanities, unknown, and broad (i.e., multiple research areas).

It is important to note that we could not find any official comprehensive documentation of Elsevier's XML schema , at the time of writing. However, our conceptualization of the variables and interpretation of the data was informed by a comprehensive review of Elsevier's websites about OA, including a list of funding agreements and details about Elsevier's Open Archive program,and Elsevier’s article submission process in the context of dedicated OA agreements [@lund; @vsnu; @Els_videos]. For validation, we matched the invoicing data provided by Elsevier to data from  the Open APC initiative.

Throughout the mostly automated data gathering and analysis process, we used tools from the Tidyverse [@tidyverse] for the R programming language [@r]. To allow for efficient data manipulation and retrieval, we imported the Crossref dump to Google BigQuery, applying the rcrossref [@rcrossref] parsers to extract relevant metadata fields. We interfaced Google BigQuery with SQL with the bigrquery client [@bigrquery]. We used crminer [@crminer] to obtain the XML files from Elsevier and the xml2 package [@xml2] to parse relevant metadata. To ensure our results are reproducible, we provided all code for obtaining the data and performing the analyses with detailed documentation as RMarkdown notebooks on GitHub (<https://github.com/njahn82/elsevier_hybrid_invoicing>).

@@

During the same period, Elsevier published `r all_oa` OA articles in total (i.e., in hybrid and full OA journals), representing `r all_oa / all_els * 100`% of the total article volume (n=`r all_els`). Looking at Figure \ref{fig:oa_volume_fig}, it is apparent that the OA article volume of hybrid journals (n=`r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(n)`; `r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(prop) * 100`%) lagged behind that of Elsevier's Open Archive program (n=`r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(n)`; `r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(prop) * 100`%) and full OA journals (n=`r filter(summary_oa_type, type == "full_oa") %>% pull(n)`; `r filter(summary_oa_type, type == "full_oa") %>% pull(prop) * 100`%), which included `r sum(mirror_jn$all_articles)` articles published in `r length(unique(mirror_jn$container_title))` mirror journals.

@@

As far as we observed, OA articles in Elsevier hybrid journals were published under two possible CC licenses--CC BY or CC BY-NC-ND. CC BY allows others to distribute, remix, adapt, and build upon the licensed work, including for commercial purposes, as long as the original author is credited. In contrast, CC BY-NC-ND is less permissive as it prohibits commercial reuse and derivatives. In the five-year period from 2015-2019, the proportion of hybrid OA articles published under a CC BY-NC-ND license marginally increased and maintained a greater share than that of CC BY (see Figure \ref{fig:oa_volume_fig}-B). Interestingly, Figure \ref{fig:oa_volume_fig}-B also reveals that hybrid journals had the highest number and proportion of openly available articles licensed under CC BY (n=`r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(n)`; `r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(prop) *100`%) compared to full OA journals  (n=`r filter(cc_by_by_type, type == "full_oa") %>% pull(n)`; `r filter(cc_by_by_type, type == "full_oa") %>% pull(prop) *100`%) and Elsevier's delayed OA program (n=`r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(n)`; `r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(prop) *100`%) . Most  delayed OA articles were provided under an Elsevier user license (Els-User), which prohibits reuse for commercial purposes.

@@

Figure \ref{fig:sponsor_sector} shows the yearly distribution of central invoiced articles by year and institutional sector. While UK-based OA invoices were mainly addressed to discipline-specific governmental and non-profit research funders, invoices to the Netherlands and Norway were issued to national academic consortia representing the higher education sector. In 2019, Elsevier also launched similar agreements in countries with lower publication output including Hungary and Poland. Besides, we found that invoiced funding bodies from the UK and the US mainly represented discipline-specific funders, while the invoiced funding bodies from other countries were focused on a broad variety of disciplines (see Table \ref{tab:subject_profile}).

@@

Table \ref{tab:sponsor_top_10} presents the top ten of `r hybrid_invoice %>% filter(!is.na(oa_sponsor_cleaned)) %>% distinct(oa_sponsor_cleaned) %>% nrow()` sponsoring bodies in our sample. Together, they accounted for around 80% of centrally invoiced APCs. The table also highlights the number of distinct journals and the share of CC BY licensed articles. Notably, UK-based research funders and the US-based Bill and Melinda Gates Foundation mainly sponsored articles published under a CC BY license. In contrast, articles funded by VSNU, the European Research Council, and Norway Institutes had a much lower proportion of CC BY.

@@

ind_sponsor_df %>%
  bind_rows(overall_ind_sponsor_df) %>%
  mutate(oa_sponsor = gsub("&", "and", oa_sponsor)) %>%
  mutate_at(4:6, function(x){
    round(x*100,0)
  }) %>%
  mutate_at(5:6, function(x){
    cell_spec(x, background = make_pal(x))}) %>%
  kbl("latex", booktabs = TRUE, escape = FALSE,
      col.names = c("Hybrid OA sponsors", "Journals", "Articles", "\\%", "CC BY", "OAPC"),
      align = c("lrrrrr"),
      linesep = "",
      caption = " Top 10 Elsevier hybrid OA sponsors by centrally invoiced articles 2015-2019.  The table furthermore presents the number of Elsevier hybrid journals with at least one invoiced article and two indicators of compliance: proportion of  CC BY licensed articles  and disclosed spending through the Open APC Initiative (OAPC.") %>%
  row_spec(11:12, hline_after = TRUE) %>%
  add_header_above(c(" " = 4, "Compliance (in %)" = 2)) %>%
   kable_styling(latex_options = c("scale_down"))
```

The table also highlights the proportion of APC funding publicly disclosed through the Open APC initiative, showing higher disclosure rates for  funders with a large CC BY share. On the other hand, APCs that were invoiced as part of transformative agreements were not publicly shared.

Figure \ref{fig:field_funder_matrix} shows the subject profile of OA sponsors based on to the journals’ subject fields of centrally invoiced articles. As expected, the figure shows that subject-specific funders, such as the British Engineering and Physical Sciences Research Council (EPSRC), tended to focus on a limited set of subject fields. In contrast, national consortia, such as VSNU and Norwegian Institutes (NO), financed OA articles  for a broader range of fields. However, within these consortia, the subject distribution also varied. VSNU OA spending, for example, had a strong focus on articles relating to medicine, environmental science, and energy, presumably reflecting disciplines with a generally strong publication output at Dutch universities. Norwegian universities mostly published hybrid OA in engineering, environmental science and agricultural and biological sciences journals.
